---
title: "Cleaning Scottish Census Data"
author: "Dave"
format: html
editor: visual
execute:
  output: false
---

# Cleaning Scottish Census Data

## Housekeeping

### Libraries

```{r libs}
library(here)
library(fs)
library(vroom)
library(dplyr)
library(readr) # for parse_number
library(janitor) # for clean_names
library(stringr)
library(tidyr)
```

### File paths

```{r paths}
in_csvs <- function(...) here("NRS-bulk-tables", ...)
```

## Cleaning Data

## Helpers

### 1. Correctly name first column

The first column of each table is output area codes but it had no name

```{r makeNameCodeCol}
name_code_col <- function(x) rename(x, oa_code = "x1")
```

### 2. Remove `all_people` or `all_households` column

```{r removeAllPeople}
remove_all_people <- function(x) select(x, -all_people)
remove_all_households <- function(x) select(x, -all_households)
```

### 3. Remove totals row

```{r removetotalsRow}
remove_totals_row <- function(x) filter(x, oa_code != "S92000003")
```

### 4. Parse numbers

Apparently NRS also put commas into their numbers? I don't know man. These are the bulk downloads for 'putting on to your own system'. smh

Oh also they code zeroes as a hypen? Let's fix that in the same process

```{r parseNumbers}
parse_numbers <- function(x, not = oa_code){
  not = enexpr(not)
  out <- mutate(
    x, 
    across(-!!not & where(is.character),
           \(x){
             replace(x, x == "-", 0) |> 
               parse_number()
           }))
  return(out)
}
```

### Big helper wrapper

Combine several of the helpers into one function

```{r generalClean}
general_clean <- function(x, remove_all = "people"){
  
  stopifnot(remove_all %in% c("people", "households"))
  if(remove_all == "people") remover <- remove_all_people
  else remover <- remove_all_households
  
  clean_names(x) |> 
    name_code_col() |> 
    remover() |> 
    remove_totals_row() |> 
    parse_numbers()
}
```

### Cleaning Age Data

```{r cleaningAgeData}
age <- vroom(in_csvs("SAPE_SYOA_ALL_2021.csv"),
                   guess_max = Inf, show_col_types = FALSE)

age <- filter(age, FeatureType == "2011 Data Zone",
              Measurement == "Count",
              Units == "People",
              Sex == "All")

age <- select(age, dz_2011_code = FeatureCode, Age, Value) 

age <- pivot_wider(age, names_from = "Age", values_from = "Value")

age <- clean_names(age)

age <- rename_with(age, str_replace, -dz_2011_code, "x", "age_")

# age <- general_clean(age)
# 
# age <- select(age, -mean_age, -median_age)
# 
# ## add age_suffix (req. 9)
# age <- rename_with(age,
#                    str_replace,
#                    where(is.numeric),
#                    "(?<=^)x", "age_")
# ## age years_suffix (req. 12)
# age <- rename_with(age,
#                    str_c,
#                    where(is.numeric) & !age_90_and_over,
#                    "_years")
# ## remove "to_" (req. 10)
# age <- rename_with(age,
#                    str_remove,
#                    where(is.numeric), 
#                    "_to")
# ## Rename final var name
# age <- rename(age, age_90_years_and_over = "age_90_and_over")
```

### Cleaning ethnicity data

```{r cleanEthnicity}
ethnicity <- vroom(in_csvs("KS201SC.csv"),
                   guess_max = Inf, show_col_types = FALSE)

ethnicity <- general_clean(ethnicity)
```

### Cleaning religion data

```{r cleanReligion}
religion <- vroom(in_csvs("KS209SCb.csv"),
                   guess_max = Inf, show_col_types = FALSE)

religion <- general_clean(religion)
```

### Cleaning household composition data

```{r cleaningHouseholdCompositionData}
household_composition <- vroom(in_csvs("KS105SC.csv"),
                   guess_max = Inf, show_col_types = FALSE)

household_composition <- general_clean(household_composition,
                                       remove_all = "households")
```

### Cleaning general health data

```{r cleaningGeneralHealth}
general_health <- vroom(in_csvs("KS301SCb.csv"),
                   guess_max = Inf, show_col_types = FALSE)

general_health <- general_clean(general_health)

general_health <- select(general_health, oa_code, starts_with("general_health"))
general_health <- rename_with(
  general_health, 
  str_remove,
  where(is.numeric),
  "general_health_"
)
```

## Save all data

```{r save_census_data}

census_dir <- function(...){
  here("census-data_scotland", ...)
}

save_census_data <- function(x, name, geography = "oa", year = "2011"){
  full_name <- str_c("scotland", name, geography, year, sep = "_")
  
  rds_name <- str_c(full_name, ".RDS")
  csv_name <- str_c(full_name, ".csv")
  
  vroom_write(x, census_dir(csv_name), delim = ",")
  saveRDS(x, census_dir(rds_name))
}

save_census_data(age, "age", "dz", "2021")
save_census_data(ethnicity, "ethnicity")
save_census_data(general_health, "general-health")
save_census_data(household_composition, "household-composition")
save_census_data(religion, "religion")
```
